The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-16 kalman_walk_paper.tex
added Haldane's rule. edited the discussion.

% Other title ideas:
%
% Rapid speciation despite conservation of phenotype
%
% How fast does network drift create incompatibilities?
%
% More than one way to grow a cat: speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network drift and speciation
%
% More than one way to grow a cat: regulatory network redundancy and speciation
%
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
%
% More than one way to grow a cat: can network drift lead to speciation?
%
% Beyond the snowball: a quantitative model of incompatibility accumulation
%
% Systems drift and speciation: more than one way to grow a cat
%
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
%
% Evolutionary conservation of phenotype does not entail conservation of the underlying
% molecular mechanism leading to rapid speciation
%
% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
%
% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
% phenotypic conservation
%
% Hybrid incompatibilities can evolve rapidly due to developmental systems drift
%
% Evolutionary systems theory and speciation

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
these concepts can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental systems drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse,  dalal2016transcriptional, dalal2017transcription}.
% are these good here? {taylor2016diverse, matsui2015regulatory, stergachis2014conservation}.

EDIT IN THESE REFS?
The literature is filled with detailed observations of molecular systems and their diversity. 
\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
There are examples of significant diversity in the networks underlying processes such as 
circadian rhythm \citep{sancar2008intelligent}, 
cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
pattern formation \plr{cite?}, 
and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.

@@

In this paper we outline a theoretical framework to study the evolution of biological systems
that can be described by systems of differential equations, such as gene regulatory networks.
We study the scenario
where the optimal phenotype remains constant over evolutionary time,
so that despite strong stabilizing selection for phenotype,
neutral drift in the underlying genotype is possible.
Results from systems theory provide
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
which gives concrete expectations for how any such
system can, in principal, undergo systems drift and rewire. 
Even with stabilization selection, a population will explore the set of all possible phenotypically equivalent gene networks.
Consequentially, two populations isolated for a sufficiently long period of time, 
will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

@@

While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
%, although it is at present still unclear the relative contribution of each process. 
System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.

Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. It also seems reasonable that systems drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, and that (5) gene misregulation can cause dysfunction (by definition).

@@

@@

% nonlinearity
What about nonlinearity?
As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}. As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations. Most notable is linearity. It is often stated that life is not linear. 
This is often true, 
%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory. Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.

>>>>

% Other title ideas:
%
% Rapid speciation despite conservation of phenotype
%
% How fast does network drift create incompatibilities?
%
% More than one way to grow a cat: speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network drift and speciation
%
% More than one way to grow a cat: regulatory network redundancy and speciation
%
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
%
% More than one way to grow a cat: can network drift lead to speciation?
%
% Beyond the snowball: a quantitative model of incompatibility accumulation
%
% System drift and speciation: more than one way to grow a cat
%
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
%
% Evolutionary conservation of phenotype does not entail conservation of the underlying
% molecular mechanism leading to rapid speciation
%
% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
%
% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
% phenotypic conservation
%
% Hybrid incompatibilities can evolve rapidly due to developmental system drift
%
% Evolutionary systems theory and speciation

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
these concepts can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse,  dalal2016transcriptional, dalal2017transcription}.
% are these good here? {taylor2016diverse, matsui2015regulatory, stergachis2014conservation}.

EDIT IN THESE REFS?
The literature is filled with detailed observations of molecular systems and their diversity. 
\plr{Diversity doesn't imply system drift -- only if the diverse systems are homologous.}
There are examples of significant diversity in the networks underlying processes such as 
circadian rhythm \citep{sancar2008intelligent}, 
cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
pattern formation \plr{cite?}, 
and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}.

@@

In this paper we outline a theoretical framework to study the evolution of biological systems
that can be described by systems of differential equations, such as gene regulatory networks.
We study the scenario
where the optimal phenotype remains constant over evolutionary time,
so that despite strong stabilizing selection for phenotype,
neutral drift in the underlying genotype is possible.
Results from systems theory provide
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
which gives concrete expectations for how any such
system can, in principal, undergo system drift and rewire. 
Even with stabilization selection, a population will explore the set of all possible phenotypically equivalent gene networks.
Consequentially, two populations isolated for a sufficiently long period of time, 
will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

% nonlinearity
%What about nonlinearity?
As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}. As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations. Most notable is linearity. It is often stated that life is not linear. 
This is often true, 
%however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). 
however, we see this as a necessary first step in the direction of a more life-like (nonlinear, perhaps) evolutionary systems theory. Depending on an actual biological system's particularities, its non-linearity, may buffer or exacerbate effects elucidated in this paper.

@@

@@

While sumarizing speciation research \citet{orr2004speciation} remarked that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
%, although it is at present still unclear the relative contribution of each process. 
System drift induced hybrid breakdown occurs rapidly even in the absence of postive Darwinian selection, suggesting that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.

Despite the results of this inquiry, in practice it remains unclear whether system drift or other speciation mechanisms predominate, as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. It also seems reasonable that system drift could be common as there is evidence (1) that the set of molecular mechanisms underlying a trait is degenerate (illustrated in this paper), (2) that there is significant regulatory variation segregating within populations, (3) that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, (4) transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that (5) gene misregulation can cause dysfunction (by definition), and (6) that numerous cryptic species complexes exist (e.g. sun skinks \citep{barley2013challenge}).

@@
